% بھائی-lang — research paper
% Compile with: xelatex bhai.tex && bibtex bhai && xelatex bhai.tex && xelatex bhai.tex
% Requires: XeLaTeX, polyglossia, fontspec, an Urdu Nastaliq font installed.
% Tested with Noto Nastaliq Urdu (free, install from Google Noto).

\documentclass[10pt,twocolumn]{article}

% ─── Multi-script (English + Urdu) ───────────────────────────────────────
\usepackage{fontspec}
\usepackage{polyglossia}
\setdefaultlanguage{english}
\setotherlanguage{urdu}
\newfontfamily\urdufont[Script=Arabic, Scale=1.05]{Noto Nastaliq Urdu}
\newcommand{\ur}[1]{{\urdufont #1}}

% ─── Layout ──────────────────────────────────────────────────────────────
\usepackage[margin=0.85in]{geometry}
\usepackage{microtype}
\usepackage{xcolor}
\usepackage{hyperref}
\hypersetup{colorlinks=true, linkcolor=black, citecolor=blue!60!black, urlcolor=blue!60!black}

% ─── Code listings ───────────────────────────────────────────────────────
\usepackage{listings}
\definecolor{kw}{RGB}{20,90,180}
\definecolor{cmt}{RGB}{100,120,100}
\definecolor{str}{RGB}{160,60,30}
\lstset{
  basicstyle=\ttfamily\footnotesize,
  commentstyle=\color{cmt}\itshape,
  keywordstyle=\color{kw}\bfseries,
  stringstyle=\color{str},
  showstringspaces=false,
  frame=single,
  framesep=4pt,
  columns=fullflexible,
  breaklines=true,
  xleftmargin=4pt,
  xrightmargin=4pt,
}

% ─── Bibliography ────────────────────────────────────────────────────────
\usepackage[backend=bibtex,style=numeric,sorting=none]{biblatex}
\addbibresource{refs.bib}

% ─── Title ───────────────────────────────────────────────────────────────
\title{
  \LARGE \ur{بھائی}-lang: A Kinship-Typed Programming Language\\
  for Cryptographically Verifiable Data Lineage
  \vspace{4pt}
}
\author{
  Adil \\
  \small Karachi
}
\date{}

\begin{document}
\twocolumn[\maketitle
\begin{abstract}
\noindent Conventional data-lineage tools (OpenLineage, DataHub, Marquez) treat provenance as an annotation layer over heterogeneous systems and recognise at most three coarse edge types: \textsc{derived\_from}, \textsc{depends\_on}, and \textsc{contains}. We argue this vocabulary is structurally insufficient for the lineage questions modern compliance, ML governance, and AI auditability actually demand.

We present \ur{بھائی}-lang, a programming language in which data lineage is the runtime substrate rather than a post-hoc instrumentation layer. Every value is a \emph{\ur{کردار}} (\emph{kirdaar} — character) carrying its ancestry; every primitive operation extends a typed kinship graph in which $\sim$50 distinct edges encode relationships such as \ur{سوتن}~(\emph{sautan}: fallback parent), \ur{سمدھی}~(\emph{samdhi}: cross-table-join co-source), \ur{استاد}~(\emph{ustaad}: training-data $\to$ model), and \ur{گواہ}~(\emph{gawah}: audit observer). The vocabulary is drawn directly from Urdu kinship terms, for which English-derived data-systems language has no equivalents.

Every \ur{کردار} is content-addressable via a Merkle-chained SHA-256 we call \ur{گواہی} (\emph{gawahi} — testimony). The cryptographic structure makes lineage \emph{verifiable} rather than vendor-asserted: a recipient of a serialised lineage graph can confirm, without trusting the producer, that no value or edge has been altered.

We describe the language design, type system, cryptographic protocol, and a $\sim$1{,}500-line reference implementation. We illustrate the approach on four case studies: GDPR delete cascade, ML feature PII leak detection, JOIN provenance via a kinship-typed merge, and tamper detection on a serialised audit. We argue that lineage-native runtime semantics are technically necessary, vocabulary-rich kinship typing is conceptually overdue, and Merkle-chained provenance is procurable.
\end{abstract}
\vspace{12pt}]

% ═════════════════════════════════════════════════════════════════════
\section{Introduction}

Data-lineage failures are expensive. The European Union has issued more than \euro{}4.5\,billion in GDPR penalties since 2018, a substantial fraction tied to inadequate ability to identify and delete derived data on request~\cite{gdpr-fines}. ML governance teams cannot reliably answer ``did sensitive feature $X$ contribute to this prediction?''~\cite{mlflow-lineage,featurestore}. AI agents call other agents in nondeterministic chains whose provenance disappears within hours~\cite{ai-agent-provenance}.

The dominant industry response is bolt-on instrumentation. Tools such as OpenLineage~\cite{openlineage}, DataHub~\cite{datahub}, and Marquez~\cite{marquez} expose APIs that data-producing systems may call to record lineage events. They are voluntary, eventually consistent, and — critically — operate over a vocabulary of three relationships: \textsc{derived\_from}, \textsc{depends\_on}, and \textsc{contains}. This impoverished type system cannot distinguish

\begin{itemize}
\item a value derived from training data versus one derived from inference inputs;
\item a join between two source tables versus an inheritance from a single one;
\item a fallback computation that consulted a backup source versus a primary path;
\item an externally imported value versus a locally derived one;
\item a tampered ancestor versus a legitimate one.
\end{itemize}

Yet these are exactly the distinctions that compliance, ML, and security teams are paid to enforce. The vocabulary gap is not a missing feature; it is structural.

\paragraph{Insight.}
Human kinship languages have $\sim$50 distinct relationship terms because we needed them. Cousin-via-paternal-uncle (\ur{چچا‌زاد}) is computationally distinct from cousin-via-maternal-aunt (\ur{خالہ‌زاد}); co-wife (\ur{سوتن}) is computationally distinct from sister (\ur{بہن}); witness (\ur{گواہ}) is computationally distinct from guarantor (\ur{ضامن}). Karachi Urdu, as a melting-pot dialect, encodes one of the world's richest kinship vocabularies. Mapping this vocabulary onto data lineage yields exactly the relationship distinctions data systems have always lacked names for.

\paragraph{Contributions.}
\begin{enumerate}
\item A programming-language design (\ur{بھائی}-lang) in which lineage is the runtime substrate rather than an annotation layer (\S\ref{sec:design}).
\item A kinship-typed lineage edge ontology of $\sim$50 distinct relationships, with formal computational semantics for the core sub\-set (\S\ref{sec:typesystem}).
\item A cryptographic testimony protocol (\ur{گواہی}) that makes lineage Merkle-chain verifiable, with $O(N)$ verification cost (\S\ref{sec:crypto}).
\item A reference implementation in $\sim$1{,}500 lines of Python and four case studies (\S\ref{sec:impl}, \S\ref{sec:cases}).
\end{enumerate}

% ═════════════════════════════════════════════════════════════════════
\section{Background and the Vocabulary Gap}
\label{sec:background}

\subsection{The PROV-DM Lineage}

The W3C PROV data model~\cite{prov-dm} formalises provenance as a tripartite graph of \emph{Entities}, \emph{Activities}, and \emph{Agents}, connected by edges including \texttt{wasGeneratedBy}, \texttt{wasDerivedFrom}, \texttt{wasAssociatedWith}, and \texttt{actedOnBehalfOf}. PROV-DM is influential and well-specified, but its edge alphabet is small and operates at the granularity of Activities (jobs, runs) rather than individual values. Implementations such as OpenLineage~\cite{openlineage} preserve the PROV-DM granularity but reduce the edge vocabulary further, typically to a single \textsc{output\_of} relation.

\subsection{Capability and Information-Flow Languages}

Capability-secure languages (E~\cite{e-language}, Pony~\cite{pony}, Monte~\cite{monte}) carry authority alongside values, ensuring that holding a reference confers the right to use it. Information-flow type systems (FlowCaml~\cite{flowcaml}, Jif~\cite{jif}, LIO~\cite{lio}) propagate confidentiality labels through computation. Both bodies of work bear on lineage but treat it as a property of \emph{access} or \emph{label}, not of derivation \emph{relationship}. They give us secrecy bits, not kinship structure.

\subsection{ML Lineage Tools}

Feast~\cite{feast}, Tecton~\cite{tecton}, MLflow~\cite{mlflow}, and Weights~\&~Biases~\cite{wandb} record feature and experiment lineage. Their semantics are nominal — every artefact has a name and a parent set — but the edges are typed only as \textsc{produced} or \textsc{used}. ``Did this feature derive from sensitive data?'' is, in 2025, still answered by manual code audit at every team we have spoken to.

\subsection{Verifiable Provenance}

Git~\cite{git} pioneered the use of a Merkle DAG to make commit history tamper-evident. Certificate Transparency~\cite{ct-rfc} extends the idea to TLS certificate logs. Datalog-with-provenance research~\cite{provenance-semirings} expresses lineage as a semiring annotation on relational query results. None of this work has, to our knowledge, been carried into a runtime where individual program values are themselves Merkle-chained.

\subsection{The Vocabulary Gap, Quantified}

Table~\ref{tab:vocab-gap} compares the edge-type expressivity of representative tools.

\begin{table}[h]
\centering
\small
\begin{tabular}{lr}
\hline
System & Distinct edge types \\
\hline
OpenLineage~\cite{openlineage} & 3 \\
DataHub~\cite{datahub} & 4 \\
Marquez~\cite{marquez} & 3 \\
PROV-DM~\cite{prov-dm} & 13 \\
Atlan~\cite{atlan} & 5 \\
\textbf{\ur{بھائی}-lang (this work)} & \textbf{$\sim$50} \\
\hline
\end{tabular}
\caption{Edge-type expressivity across data-lineage systems. PROV-DM is the formal academic baseline; commercial tools have collapsed it further. \ur{بھائی}-lang ships with $\sim$50 typed edges drawn from Urdu kinship vocabulary.}
\label{tab:vocab-gap}
\end{table}

% ═════════════════════════════════════════════════════════════════════
\section{Design: Lineage as Runtime Substrate}
\label{sec:design}

\subsection{The \ur{کردار} Value Model}

Every value in \ur{بھائی} is a \ur{کردار} (kirdaar) — a record carrying:

\begin{itemize}
\item a primitive value (number, string, boolean, list, null);
\item a set of typed parent edges $\{(r_i, k_i)\}$, each $r_i \in \mathsf{Rel}$ a relation label, $k_i$ another \ur{کردار};
\item a consent state $\in \{\textsc{vip}, \textsc{do-numri}, \textsc{shopping}\}$;
\item an optional trust label $\in \{\ur{جانی}, \ur{بھائی}, \bot\}$;
\item an optional sensitivity label $\in \{\ur{حساس}, \bot\}$;
\item an adversarial bit \ur{دشمن} $\in \{\top, \bot\}$;
\item an origin marker $\in \{\ur{لے_پالک}, \ur{رضاعی}, \bot\}$.
\end{itemize}

\subsection{Operations and Edge Construction}

Every primitive operation \emph{constructs} edges. Binary arithmetic \texttt{c = a + b} yields a fresh \ur{کردار} $c$ with $\mathsf{parents}(c) = \{(\ur{باپ}, a), (\ur{ماں}, b)\}$ — left operand as paternal lineage, right operand as maternal. Higher-arity operations (joins, training declarations) construct correspondingly typed edges.

There is no opt-in. Lineage is created \emph{by construction} as a side effect of computation; it cannot be omitted, forgotten, or buggy.

\subsection{Propagation Rules}

The runtime enforces three propagation laws on every operation, illustrated for the binary case:
\begin{align*}
\mathsf{trust}(c)       &= \max(\mathsf{trust}(a), \mathsf{trust}(b)) \\
\mathsf{sens}(c)        &= \mathsf{sens}(a) \vee \mathsf{sens}(b) \\
\mathsf{dushman}(c)     &= \mathsf{dushman}(a) \vee \mathsf{dushman}(b) \\
\mathsf{consent}(c)     &= \begin{cases}
  \textsc{do-numri} & \text{if } a \text{ or } b \notin \textsc{vip} \\
  \textsc{vip} & \text{otherwise}
\end{cases}
\end{align*}
Trust is monotone under join, sensitivity and adversarial flags propagate as taint, and consent demotes upon any compromised ancestor.

% ═════════════════════════════════════════════════════════════════════
\section{The \ur{رشتہ} Type System}
\label{sec:typesystem}

\subsection{Edge Vocabulary}

Table~\ref{tab:edges} lists the core typed edges and their computational meaning. The full ontology contains $\sim$50 entries derived from the kinship structure of Urdu — including paternal-versus-maternal nephew (\ur{بھتیجا} vs \ur{بھانجا}), co-brothers-in-law via married sisters (\ur{ہم‌زلف}), milk-kin (\ur{رضاعی}), and adopted (\ur{لے‌پالک}) — each with a distinct semantic.

\begin{table*}[t]
\centering
\small
\begin{tabular}{lll}
\hline
Edge & Literal Urdu meaning & Computational semantic \\
\hline
\ur{باپ} (baap) & father & primary derivation (left operand) \\
\ur{ماں} (maa) & mother & primary derivation (right operand) \\
\ur{استاد} (ustaad) & teacher & training-data $\to$ trained-model edge (ML governance) \\
\ur{شاگرد} (shagird) & apprentice & inverse of \ur{استاد}: a model is a \ur{شاگرد} of its training set \\
\ur{سوتن} (sautan) & co-wife & fallback parent: alternative derivation when primary is unavailable \\
\ur{گواہ} (gawah) & witness & read-only audit snapshot — sees the value but does not consume \\
\ur{بیوی}/\ur{شوہر} & wife / husband & a JOIN: two source values merged into a child \\
\ur{سمدھی} (samdhi) & co-in-laws & two source values that share a JOIN child (cross-table relationship) \\
\ur{ہم‌زلف} (ham-zulf) & co-brothers-in-law & parallel pipelines sharing a common JOIN partner \\
\ur{بھتیجا} (bhateeja) & paternal nephew & value derived from a sibling-of-paternal-lineage \\
\ur{بھانجا} (bhanja) & maternal nephew & value derived from a sibling-of-maternal-lineage \\
\ur{چچازاد} (chacha-zaad) & paternal cousin & shares a paternal grandparent — same raw data, different pipeline \\
\ur{رضاعی} (razai) & milk-kin & cached / mirrored copy: same content, separate identity \\
\ur{لے‌پالک} (le-paalak) & adopted & externally imported — no native lineage in this run \\
\ur{دشمن} (dushman) & enemy & adversarial mark: do not derive from this without quarantine \\
\hline
\end{tabular}
\caption{Core typed lineage edges in \ur{بھائی}-lang and their computational semantics. Each edge is a Karachi-Urdu kinship term whose social meaning maps directly onto a data-systems relationship that conventional vocabulary cannot name.}
\label{tab:edges}
\end{table*}

\subsection{Why Vocabulary Matters}

Three edges in particular have no name in any data-systems tool we surveyed:

\paragraph{\ur{سوتن} (co-wife) as fallback parent.}
A common pattern in resilient computation is: derive a value from a primary source, but fall through to an alternative if the primary is unavailable. SQL implements this as \texttt{COALESCE}; ML pipelines implement it as model ensembles or A/B tests. In each case, the \emph{relationship} between the two parents and the child is absent from the lineage graph. \ur{بھائی} encodes it as a \ur{سوتن} edge — declaratively, an alternative parent of equal lineage standing.

\paragraph{\ur{سمدھی} (co-in-laws) as JOIN-induced relation.}
When tables \texttt{users} and \texttt{orders} are joined, the rows of the result are children of both. The relationship between \texttt{users} and \texttt{orders} themselves — they have shared descendants — is what kinship terms \emph{co-in-laws}. Every JOIN in a data warehouse creates a \ur{سمدھی} relationship; no lineage tool surfaces it. \ur{بھائی} does so by construction.

\paragraph{\ur{استاد} (teacher) as training edge.}
The training-data-to-model edge is the most-asked-about relationship in AI governance and the least-typed in lineage tools. \ur{بھائی} represents it directly: a model is a \ur{شاگرد} of its training data; the \texttt{\ur{استاد\_چین} <model>} query walks only \ur{استاد} edges, yielding the complete training provenance.

% ═════════════════════════════════════════════════════════════════════
\section{Cryptographic Testimony (\ur{گواہی})}
\label{sec:crypto}

\subsection{Threat Model}

We assume:
\begin{itemize}
\item the producer of a serialised lineage graph may be adversarial;
\item the recipient (auditor, regulator, downstream consumer) wishes to verify the lineage without trusting the producer;
\item the recipient has the public hash function but no shared secret;
\item integrity, not confidentiality, is the property of interest.
\end{itemize}
This is the standard threat model for tamper-evident logs.

\subsection{Hash Construction}

For every \ur{کردار} $k$ we define
\[
\mathsf{gawahi}(k) = H\!\left( \mathsf{enc}(k.v, k.\tau, P_k) \right)
\]
where $H$ is SHA-256, $k.v$ is the primitive value, $k.\tau$ is the tuple of consent / trust / sensitivity / adversarial / origin tags, and
\[
P_k = \mathsf{sort}\bigl( \{(r_i, \mathsf{gawahi}(k_i)) : (r_i, k_i) \in \mathsf{parents}(k)\} \bigr).
\]
The encoding $\mathsf{enc}$ is canonical JSON. Because $\mathsf{gawahi}$ is defined recursively over parent hashes, the resulting structure is a Merkle DAG isomorphic to the lineage graph itself.

\subsection{Properties}

\paragraph{Determinism.} Two \ur{کردار} with identical values, tags, and parent hashes have identical \ur{گواہی}, regardless of name, allocation order, or session.

\paragraph{Tamper-evidence.} Modifying any field of any \ur{کردار} alters its hash and, by the recursion, the hash of every descendant. A single-bit change in a root \ur{کردار} invalidates every \ur{گواہی} below it.

\paragraph{Verification cost.} Linear in the number of \ur{کردار}: each is hashed once, each parent edge contributes a constant-size string. For the audit graphs we construct in \S\ref{sec:cases} (10--100 nodes), verification completes in milliseconds.

\paragraph{What it does \emph{not} catch.} A producer can elide \ur{کردار} entirely from the serialised graph — the hash of the omitted node is simply absent. \ur{بھائی} addresses elision orthogonally via globals tracking, but a determined producer with no external commitment can still omit. We discuss external anchoring in~\S\ref{sec:discussion}.

\subsection{Round-trip Verification}

The \texttt{\ur{سنبھال}} primitive emits, alongside each \ur{کردار}, its \ur{گواہی}. The \texttt{\ur{اٹھا}} primitive recomputes hashes after reconstructing the graph and aborts with the offending name on mismatch. We have validated this end-to-end on a graph of 5 \ur{کردار}, including a manual single-byte tamper test.

% ═════════════════════════════════════════════════════════════════════
\section{Implementation}
\label{sec:impl}

The reference implementation is two tree-walking interpreters in $\sim$1{,}500 lines of Python:

\begin{itemize}
\item \texttt{bhai.py} — a general-purpose language with closures, control flow, lists, and a 19-builtin standard library (file IO, time, math, regex). Useful for writing real programs that interact with the operating system.
\item \texttt{rishta.py} — the lineage-native DSL described in this paper. No control flow; programs are sequences of declarations and queries against the lineage graph.
\end{itemize}

Both interpreters share the same lexer architecture (Urdu identifier handling, Urdu digit literals \ur{۰}--\ur{۹}, Urdu punctuation \ur{،}~/~\ur{۔}). They support modules via \texttt{\ur{منگوا}} (cycle-detected import). The runtime is a graph of \ur{کردار} objects with cached-on-demand hash computation. Persistence emits canonical JSON; load verifies hashes before exposing globals.

The implementation is open-source~\cite{bhai-repo} and runs on Python 3.8+ with no third-party dependencies.

% ═════════════════════════════════════════════════════════════════════
\section{Case Studies}
\label{sec:cases}

\subsection{GDPR Delete Cascade}

\begin{lstlisting}[caption={GDPR cascade in 7 lines.},captionpos=b]
کردار صارف = "ali@karachi.pk" حساس جانی
کردار پروفائل = صارف + " (کراچی)"
کردار رپورٹ = پروفائل + " — اپریل 2026"

تپکا صارف
# ⚠ صارف تپکا — 4 اولاد دو نمبری ہوئی
پھوٹ رپورٹ
# رپورٹ: (ہائیڈ) [دو نمبری] [جانی] [حساس]
\end{lstlisting}

A single \ur{تپکا} (revoke) call cascades through every value derived from \ur{صارف}, marking each \ur{دو\,نمبری} (tainted). Reads of tainted \ur{کردار} return a hidden sentinel rather than the underlying value. The compliance team's three-week audit becomes a four-token statement.

\subsection{ML Feature PII Leak Detection}

\begin{lstlisting}[caption={Feature provenance audit.},captionpos=b]
کردار سی_این_آئی_سی = 4210112345678 حساس
کردار عمر = 25
کردار ماڈل_فیچر = سی_این_آئی_سی + عمر
رساؤ ماڈل_فیچر
# ⚠ حساس رساؤ — 2 حساس جد ملے:
#   ← خود: ماڈل_فیچر [VIP] [حساس]
#   ← باپ: سی_این_آئی_سی [VIP] [حساس]
\end{lstlisting}

The \ur{رساؤ} (leak) query walks ancestors filtered by sensitivity. Any feature whose lineage includes a \ur{حساس} \ur{کردار} is flagged. This addresses the canonical question of ML governance — \emph{did sensitive data influence this prediction?} — by lineage walk rather than code review.

\subsection{JOIN Provenance via \ur{سمدھی}}

\begin{lstlisting}[caption={Cross-table JOIN provenance.},captionpos=b]
کردار صارفین = 100 حساس
کردار آرڈرز = 50
جوڑ صارفین سے آرڈرز بنا فل_ٹیبل

سمدھی صارفین
#   ↔ آرڈرز = 50 [VIP] (via فل_ٹیبل)
\end{lstlisting}

The \ur{جوڑ} (join) primitive constructs \ur{بیوی} / \ur{شوہر} edges from the result to each source. The induced \ur{سمدھی} relationship between sources is queryable on demand. This is the cross-table ``what touches what'' question every BI team computes ad hoc; here it is graph-intrinsic.

\subsection{Tamper Detection}

We construct a 5-node graph, save it via \ur{سنبھال}, and externally modify the value of one \ur{کردار} in the resulting JSON. The subsequent \ur{اٹھا} aborts:
\begin{quote}\small
\texttt{⚠~\ur{گواہی نا کام: 'صارف' میں چھیڑ چھاڑ ہوئی!}\\
expected \texttt{a6b3b9ea073d}\ldots, file says \texttt{e169a0b214be}\ldots}
\end{quote}
The hash chain immediately localises tampering to the responsible \ur{کردار}.

% ═════════════════════════════════════════════════════════════════════
\section{Discussion}
\label{sec:discussion}

\subsection{What This Buys}

The combination of (a) lineage-by-construction, (b) kinship-typed edges, and (c) cryptographic testimony eliminates three classes of bug at once: lineage absence, lineage misclassification, and lineage tampering. Compliance audits that today require multi-person quarter-long efforts collapse to graph queries that complete in milliseconds and produce machine-verifiable artefacts.

\subsection{Limitations}

\paragraph{Performance.} The reference interpreter is tree-walking. Realistic workloads will require a bytecode VM and incremental hash maintenance.

\paragraph{Mutation.} The current language is value-immutable. A versioned-mutation primitive (\ur{سافٹویئر اپڈیٹ}) is on the roadmap.

\paragraph{Elision.} Cryptographic testimony catches modification but not omission. External anchoring (commitment to a public log such as Certificate Transparency or a blockchain) closes this gap; the integration is straightforward but unimplemented.

\paragraph{Adoption.} The Urdu-script keyword choice is a cultural commitment. We expect a productionised English-keyword variant will be required for enterprise adoption; the underlying semantics transfer unchanged.

\subsection{Cultural Framing}

The decision to express the type system in Urdu kinship terms is deliberate and not decorative. Karachi Urdu's vocabulary is the most relationship-rich substrate we found; the language design is downstream of the vocabulary, not upstream of it. We conjecture that other natural-language kinship systems (Tamil, Mandarin, Yoruba) would yield comparable insight, and that programming-language design has historically been impoverished by drawing only from English-derived ontologies.

\subsection{Future Work}

Four directions are immediate: (i) external anchoring of \ur{گواہی} to public logs; (ii) inter-organisation lineage-passport format for verifiable hand-off between data producers and consumers; (iii) compile-to-Python embedding so existing codebases can opt in incrementally; (iv) more typed edges from the kinship vocabulary not yet exposed.

% ═════════════════════════════════════════════════════════════════════
\section{Related Work}

Beyond the work surveyed in \S\ref{sec:background}: PASS~\cite{pass} pioneered OS-level provenance capture; Burrito~\cite{burrito} captured scientific-workflow provenance; Carmine~\cite{carmine} explored provenance for distributed dataflow. None typed provenance edges beyond a small fixed alphabet, and none made provenance cryptographically verifiable at the value level.

In the cryptographic-log space, Trillian~\cite{trillian} is the production substrate behind Certificate Transparency and presents a general append-only Merkle log. Our \ur{گواہی} differs in being structured over the lineage DAG itself, not over an opaque event sequence.

In programming-language design, work on effect systems~\cite{koka,frank} and capability languages most closely parallels ours in treating a meta-property of values (effects, authority) as part of the type system. \ur{بھائی} extends this tradition to lineage, with the distinguishing feature that the relationship — not just the property — is typed.

% ═════════════════════════════════════════════════════════════════════
\section{Conclusion}

We have argued that data-lineage tools have been built on an impoverished vocabulary of three to thirteen edge types, and that this vocabulary gap is responsible for a wide class of unanswerable compliance, ML, and AI auditability questions. We presented \ur{بھائی}-lang, a programming language in which lineage is the runtime substrate, edges are typed from a $\sim$50-term Karachi-Urdu kinship vocabulary, and every value is content-addressable through a Merkle-chained SHA-256 we call \ur{گواہی}. The reference implementation is small ($\sim$1{,}500 lines of Python), demonstrates the approach end-to-end, and is freely available.

The contribution of this paper is not the implementation but the conceptual move: \emph{lineage is not an annotation problem; it is a vocabulary problem solved by making derivation relationships first-class}. We invite the data-systems and programming-language communities to consider what other relationship vocabularies — drawn from the world's languages, not from English alone — might similarly enrich our tools.

\printbibliography

\end{document}
